Grove: A Convergent Collaborative Structure-Editor Calculus

### 2020-11-12 grove-paper.tex
Fix placeholder references to vertexes and edges (Michael D. Adams)

\tikzset{
  %every picture/.style={framed, background rectangle/.style={draw=gray!50}},
}
\tikzset{edge style/.style={
  draw,
  %color=gray,
  font={\small\ttfamily},
  /tikz/every edge quotes/.style={
    %draw=gray!20,
    anchor=west,
    swap/.append code={
      \ifpgfarrowswap
        \pgfkeysalso{anchor=west}
      \else
        \pgfkeysalso{anchor=east}
      \fi}},
}}
\tikzset{graphs/graph style/.style={
  tree layout,
  level distance=0.5cm,
  level sep=0.5cm,
  sibling distance=0.5cm,
  sibling sep=0.1cm,
  part distance=0.1cm,
  part sep=0.1cm,
  component distance=0.1cm,
  component sep=0.1cm,
  nodes={
    draw,
    %color=gray,
    inner sep=2pt,
    rounded corners=1mm},
  edges={edge style},
}}
\tikzset{graphs/root style/.style={
 %draw=none,
 as={\textbullet$_{0}$}
}}
\tikzset{alice/.style={
  color=red!80!black,
  font={\bfseries\small},
  thick,
}}
\tikzset{bob/.style={
  color=green!60!black,
  font={\bfseries\small},
  thick,
}}
\tikzset{merge/.style={
  color=blue,
  font={\bfseries\small},
  thick,
}}
\tikzset{alice edge/.style={alice, edge style, font={\bfseries\small}}}
\tikzset{alice node/.style={alice}}
\tikzset{alice step/.style={alice}}
\tikzset{bob edge/.style={bob, edge style}, font={\bfseries\small}}
\tikzset{bob node/.style={bob}}
\tikzset{bob step/.style={bob}}
\tikzset{merge edge/.style={merge, edge style}, font={\bfseries\small}}
\tikzset{merge node/.style={merge}}
\tikzset{merge step/.style={merge,decorate,decoration={coil,amplitude=1.0pt,segment length=7.0pt,aspect=0}}}
\tikzset{star/.style={edge node={node[inner sep=0pt,at end,sloped] {\large\textbf{${}^{*}$}}}}}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% \figureDisconnection
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newedge{DisconnectionAlice}{R}
\newedge{DisconnectionBob}{L}
\newcommand{\figureDisconnection}{
\begin{figure*}
%\hfill
\hskip0.12\linewidth
%%%%%%%%
\begin{subfigure}[t]{0.21\textwidth}
\centering
\caption{
\begin{tikzpicture}[remember picture, overlay]
\path (-7.5em,0cm) node (fig-Multi-User Cycles-d-x) [gray,anchor=base west] {{\textbf{\small Fig.~\ref*{fig:Multi-User Cycles-d}}}};
%%%%
\path (0cm,0cm) node (fig-Disconnection-a) [anchor=base] {\strut};
\path [draw,->,alice step] (fig-Multi-User Cycles-d-x) to[star] (-2em, 0cm |- fig-Disconnection-a);
\end{tikzpicture}
}
\label{fig:Disconnection-a}
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "\vWrapPlus" [> "\eWrapPlus"] -> {
   "\vSimpleTimes" [> "\eNestedBob"'],
   "\vMultiCyclePlus" [> "\eMultiCycleBobPlus"]
  }
 }
};
\end{tikzpicture}
\end{subfigure}
%%%%%%%%
\hfill
%%%%%%%%
\begin{subfigure}[t]{0.21\textwidth}
\centering
\caption{
\begin{tikzpicture}[remember picture, overlay]
\path (0cm,0cm) node (fig-Disconnection-b) [anchor=base] {\strut};
\path [draw,->,alice step] (fig-Disconnection-a) to[star] (-2em, 0cm |- fig-Disconnection-b);
\end{tikzpicture}
}
\label{fig:Disconnection-b}
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "\vWrapPlus" [> "\eWrapPlus"] -> {
   "\vSimpleTimes" [> "\eNestedBob"'] -> {
    {},
    "\vMultiCyclePlus" [> "\eDisconnectionAlice", >alice edge]
   },
   {}
  }
 }
};
\end{tikzpicture}
\end{subfigure}
%%%%%%%%
\hfill
%%%%%%%%
\begin{subfigure}[t]{0.21\textwidth}
\centering
\caption{
\begin{tikzpicture}[remember picture, overlay]
\path (0cm,0cm) node (fig-Disconnection-c) [anchor=base] {\strut};
\path [draw,->,bob step] (fig-Disconnection-a) to[out=15,in=165,star] (-2em, 0cm |- fig-Disconnection-c);
\end{tikzpicture}
}
\label{fig:Disconnection-c}
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "\vWrapPlus" [> "\eWrapPlus"] -> {
   {},
   "\vMultiCyclePlus" [> "\eMultiCycleBobTimes"] -> {
    "\vSimpleTimes" [> "\eDisconnectionBob"', >bob edge],
    {}
   }
  }
 }
};
\end{tikzpicture}
\end{subfigure}
%%%%%%%%
\hfill
%%%%%%%%
\begin{subfigure}[t]{0.21\textwidth}
\centering
\caption{
\begin{tikzpicture}[remember picture, overlay]
\path (0cm,0cm) node (fig-Disconnection-d) [anchor=base] {\strut};
\path [draw,->,merge step] (fig-Disconnection-b) to[out=15,in=165] (-2em, 0cm |- fig-Disconnection-d);
\path [draw,->,merge step] (fig-Disconnection-c) to (-2em, 0cm |- fig-Disconnection-d);
\end{tikzpicture}
}
\label{fig:Disconnection-d}
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "\vWrapPlus" [> "\eWrapPlus"]
 }
};
\path (-0.5cm,-2cm) graph[graph style] { plus/"\vSimpleTimes" };
\path ( 0.5cm,-2cm) graph[graph style] { times/"\vMultiCyclePlus" };
\path[draw,->,merge edge] (times) to ["\eMultiCycleBobTimes"'anchor=north,out=-135,in=-45] (plus);
\path[draw,->,merge edge] (plus) to ["\eMultiCycleBobPlus"anchor=south,out=45,in=135] (times);
\end{tikzpicture}
\end{subfigure}
%%%%%%%%
\hfill{}
\caption{Example of Disconnection.}
\label{fig:Disconnection}
\end{figure*}
}

@@

This section introduces collaborative structure editing in Grove.
We begin in \autoref{sub:Program Representation} by covering how we use graphs to represent expressions with holes,
\autoref{sub:Single-User Actions} explores edits by a single user, Alice,
and \autoref{sub:Multi-user Interactions} explores edits by multiple users, Alice and Bob,
each of whom is editing their own instance of a program.

do not model cursors
but model later in \autoref{sub:Cursors} 
they are not involved in the communication of graph edits
though their representation is slightly subtly

@@

Program state is a directed graph with a distinguished root vertex.
%For example, \autoref{fig:A simple tree with a hole-b} shows the graph corresponding to the term in \autoref{fig:A simple tree with a hole-a}.
Each vertex represents a term and is labeled with both a globally unique identifier and a term constructor.
In \autoref{fig:A simple tree with a hole-b}, \vSimpleTimes{} and \vSimpleX{} have the constructors~\texttt{*} and \texttt{var(x)},\todo{var} respectively.
The root vertex has the distinguished identifier~0 and the constructor~\textbullet.

@@

TODO: explain cross-page arrows

@@

\todo{"user actions" and "graph edits"}
To start our examples, Alice moves her cursor to the hole in~\texttt{x * \hole} in \autoref{fig:A simple tree with a hole}
and constructs the variable~\texttt{y} as shown in \autoref{fig:Construction-b}.
This action corresponds to creating \eConstructionY{} from \vSimpleX{} at the \texttt{R} position to a newly
created vertex, \vConstructionY{}, containing the variable reference~\texttt{y}.
The resulting graph, shown in \autoref{fig:Construction-b}, represents the expression \texttt{x * y}.

@@

If a user want to move code from one place to another we have two options.
The first is to delete the code from its old position, and
reconstruct the code in its new position.
However, edges and vertices have unique identifiers in Grove 
to support the collaborative editing features we discuss next,
so reconstruction will not preserve identity of vertices and edges. 
Consequently, we include explicit repositioning actions that preserve identity.
These would be triggered by, for example, a drag and drop, or a cut and paste
(but not copy and paste, or a second paste after a cut, which do not need to
preserve identity and so would rely on reconstruction as already described).

In terms of the graph, a repositioning action involves simply deleting the edge
from the original parent then adding an edge from the new parent.
Thus the code is actually \textit{moved}, not copied.

For example, in \autoref{fig:Move} Alice continues her edits by
moving \texttt{\hole{} * y} from the left child of \texttt{+} to its right child.
This corresponds to the following sequence of graph edits, shown in \autoref{fig:Move}:
(a) delete \eWrapPlus{} leaving \vSimpleTimes{} temporarily orphaned and 
(b) add \eMoveTimes{} to \vSimpleTimes{} in the right child position of \vWrapPlus{}.

@@

\marginpar{START REV HERE}

@@

@@

Alice add~\texttt{u} as the left child of \Vertex{TODO}
and Bob changes the~\texttt{y} to a~\texttt{v}.
Before transmitting their edits to each other,
Alice thus has the graph in \autoref{fig:Editing Different Parts of the Code-a}
and Bob has the graph in \autoref{fig:Editing Different Parts of the Code-b}.

%\figureCommutativity
\begin{figure}
\centering
\begin{tikzpicture}
\path (-3cm, 0cm) node (a) [align=center]       {Original \\ Version};
\path ( 0cm, 1cm) node (b) [align=center,alice node] {Alice's  \\ Version};
\path ( 0cm,-1cm) node (c) [align=center,bob node]   {Bob's    \\ Version};
\path ( 3cm, 0cm) node (d) [align=center]       {Combined \\ Version};
\path [draw,->,alice step] (a) -- node [pos=0.7,auto,align=center] {Alice's \\ Edits} (b);
\path [draw,->,bob step]   (a) -- node [pos=0.7,auto,swap,align=center]      {Bob's \\ Edits}   (c);
\path [draw,->,merge step] (b) -- node [pos=0.3,auto] {Sync} (d);
\path [draw,->,merge step] (c) -- node [pos=0.3,auto,swap]      {Sync} (d);
\end{tikzpicture}
\caption{Commutativity of Edits.}
\label{fig:Commutativity}
\end{figure}

@@

After the edits in the previous subsection synchronize,
Alice changes the~\texttt{v} to a~\texttt{w}, which results in the graph in \autoref{fig:Editing Nested Parts of the Code-a}.
On the other hand, Bob moves the~\texttt{u * v} from the left of the plus to the right of the plus
by deleting \eMoveTimes{} and adding \Edge{19}, which results in the graph in \autoref{fig:Editing Nested Parts of the Code-b}.
Since vertex and edge identifiers are preserved by these,
both Alice's and Bob's edits can be applied to the other's graph,
which results in both Alice and Bob having the graph in \autoref{fig:Editing Nested Parts of the Code-c}.

A consequence of this design is that we allow users to edit
parts of the code that have been deleted.  For example,
Alice might receive Bob's deletion of \eMoveTimes{} before making her
edits and before receiving Bob's addition of \Edge{19}.
Her edits would then be within the code deleted by Bob.
TODO: She can restore that code using a restoration action add a fresh edge to it.

@@

Finally, we address our first kind of genuine conflict between edits.
Alice inserts the variable~\texttt{x} to the right side of the~\texttt{+}
while Bob add the variable~\texttt{y} also to the right side of the~\texttt{+}.
These results in the graphs in \autoref{fig:Multi-child-a} and \autoref{fig:Multi-child-b}, respectively.
In particular pay attention to the edges going out of the \texttt{right} index of \vWrapPlus{}.
In \autoref{fig:Multi-child-a} it is \Edge{21}, which points to the variable~\texttt{x},
and in \autoref{fig:Multi-child-b} it is \Edge{23}, which points to the variable~\texttt{y}.

We merge these by including \emph{both} edges as edges going out of the \texttt{right} index
of \vWrapPlus{}.
This results in \autoref{fig:Multi-child-c}.
The conflict between Alice's and Bob's changes is
represented by the fact that there are multiple edges going out of one index (i.e., the \texttt{right}
index of \vWrapPlus{}).
In the graph representation, we flag this conflict by highlighting its edges
in red (TODO).
In the textual representation, we flag this conflict with the notation~\texttt{\{ x | y \}}.
This is a conflict that must eventually be resolved by the user by deleting
one of the edges and editing the other to be the merged value.
For example, one could delete \Edge{23} and then edit the~\texttt{x}
to be~\texttt{x * y}.
Note that users can continue to edit the code while in this conflicted state.
It just must be resolved before the the code is run.\footnote{It
may be possible to develop evaluation models that allow these sorts of conflicts,
but that is beyond the scope of this paper.}

@@

An example of this is shown in \autoref{fig:Multi-parent}.
\autoref{fig:Multi-parent-a} shows the graph after Alice and Bob have made some more edits,
and Alice and Bob's editors are synchronized with each other.
This graph represents the expression~\texttt{(w + \hole{}) + \hole{})}.
Alice moves the variable~\texttt{w} to the left-most hole,
while Bob moves it to the right-most hole.
This results in the graphs in \autoref{fig:Multi-parent-b} and \autoref{fig:Multi-parent-c},
which represent respectively the expression~\texttt{(\hole{} + w) + \hole{}} for Alice
and the expression~\texttt{(\hole{} + \hole{}) + w} for Bob.
In both cases, these edits are achieved with one edge delete and one edge add.
Alice deletes \Edge{TODO} and adds \Edge{TODO}, while Bob deletes \Edge{TODO} and adds edge \Edge{TODO}.
As explained in Subsection~TODO:REF, deletes are idempotent so Alice and Bob both deleting \Edge{TODO} is not a problem.
However, both \Edge{TODO} (added by Alice) and edge \Edge{TODO} (added by Bob) point to the same vertex.
This generates REF:FIG, in which \Vertex{TODO} as two parents pointing to it.
We call this situation a multi-parent conflict.

@@

\subsubsection{Multi-User Cycles}
\label{sub:Multi-User Cycles}

@@

For example, consider the situation in \autoref{fig:Multi-User Cycles}.
\autoref{fig:Multi-User Cycles-a} shows the graph after Alice and Bob have made some more edits,
and Alice and Bob's editors are synchronized with each other.
This graph represents the expression~\texttt{((\hole+\hole)+(\hole*\hole))+\hole}.
Alice moves the leftmost \texttt{+} into the rightmost hold
and then the \texttt{*} underneath that.
This results in \autoref{fig:Multi-User Cycles-b},
which represents the expression~\texttt{(\hole+\hole)+((\hole*\hole)+\hole)}.
On the other hand, Bob does the same move
but puts the \texttt{+} and under the \texttt{*}.
This results in \autoref{fig:Multi-User Cycles-c},
which represents the expression~\texttt{(\hole+\hole)+((\hole+\hole)*\hole)}.

On their own, neither of these edits creates a cycle.
However, merging the edit actions of both Alice and Bob results in the graph
in \autoref{fig:Multi-User Cycles-d}, which has a cycle
between \Vertex{TODO} and \Vertex{TODO}.
This situation is akin to a merge conflict in traditional version control systems.
Resolving it requires user input, but is as simple as
either deleting \Edge{TODO} and \Edge{TODO} (thus favoring Alice's version)
or deleting \Edge{TODO} and \Edge{TODO} (thus favoring Bob's version).
Since a cycle like this that is connected to the root vertex always
contains vertexes with multiple parents that if removed would break the cycle,
we use the user interface described in \autoref{sub:Multi-parent conflicts}
to display these kinds of programs.
For example, \autoref{fig:Multi-User Cycles-d} would be displayed
as shown in TODO:FIGURE.
We leave the user interface considerations of resolving the conflict such a
cycle represents to future work.

@@

For example, consider the situation in \autoref{fig:Disconnection}.
\autoref{fig:Disconnection-a} shows the graph after Alice and Bob have made some more edits,
and Alice and Bob's editors are synchronized with each other.
This graph represents the expression~\texttt{(\hole+\hole)+(\hole*\hole)}.
Alice moves the \texttt{*} under \texttt{+}$_2$.
This results in \autoref{fig:Disconnection-b},
which represents the expression~\texttt{((\hole*\hole)+\hole)+\hole}.
On the other hand, Bob puts \texttt{+}$_2$ and under the \texttt{*}.
This results in \autoref{fig:Disconnection-c},
which represents the expression~\texttt{\hole+((\hole+\hole)*\hole)}.

When these edits are merged, the result is the graph in \autoref{fig:Disconnection-d}.
Since Alice deleted the edge from \Vertex{TODO} to \Vertex{TODO}
and Bob deleted the edge from \Vertex{TODO} to \Vertex{TODO},
when these are merged \Vertex{TODO} and \Vertex{TODO}
are completely disconnected from the root.

>>>>

\tikzset{
  %every picture/.style={framed, background rectangle/.style={draw=gray!50}},
}
\tikzset{edge style/.style={
  draw,
  %color=gray,
  font={\small\ttfamily},
  /tikz/every edge quotes/.style={
    %draw=gray!20,
    anchor=west,
    swap/.append code={
      \ifpgfarrowswap
        \pgfkeysalso{anchor=west}
      \else
        \pgfkeysalso{anchor=east}
      \fi}},
}}
\tikzset{graphs/graph style/.style={
  tree layout,
  level distance=0.5cm,
  level sep=0.5cm,
  sibling distance=0.5cm,
  sibling sep=0.1cm,
  part distance=0.1cm,
  part sep=0.1cm,
  component distance=0.1cm,
  component sep=0.1cm,
  nodes={
    draw,
    %color=gray,
    inner sep=2pt,
    rounded corners=1mm},
  edges={edge style},
}}
\tikzset{graphs/root style/.style={
 %draw=none,
 as={\textbullet$_{0}$}
}}
\tikzset{alice/.style={
  color=red!80!black,
  font={\bfseries\small},
  thick,
}}
\tikzset{bob/.style={
  color=green!60!black,
  font={\bfseries\small},
  thick,
}}
\tikzset{merge/.style={
  color=blue,
  font={\bfseries\small},
  thick,
}}
\tikzset{alice edge/.style={alice, edge style, font={\bfseries\small}}}
\tikzset{alice node/.style={alice}}
\tikzset{alice step/.style={alice}}
\tikzset{bob edge/.style={bob, edge style}, font={\bfseries\small}}
\tikzset{bob node/.style={bob}}
\tikzset{bob step/.style={bob}}
\tikzset{merge edge/.style={merge, edge style}, font={\bfseries\small}}
\tikzset{merge node/.style={merge}}
\tikzset{merge step/.style={merge,decorate,decoration={coil,amplitude=1.0pt,segment length=7.0pt,aspect=0}}}
\tikzset{star/.style={edge node={node[inner sep=0pt,at end,sloped] {\textbf{\huge${}^{*}$}}}}}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% \figureDisconnection
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newedge{DisconnectionAlice}{R}
\newedge{DisconnectionBob}{L}
\newcommand{\figureDisconnection}{
\begin{figure*}
%\hfill
\hskip0.12\linewidth
%%%%%%%%
\begin{subfigure}[t]{0.21\textwidth}
\centering
\caption{
\begin{tikzpicture}[remember picture, overlay]
\path (-7.5em,0cm) node (fig-Multi-User Cycles-d-x) [gray,anchor=base west] {{\textbf{\small Fig.~\ref*{fig:Multi-User Cycles-d}}}};
%%%%
\path (0cm,0cm) node (fig-Disconnection-a) [anchor=base] {\strut};
\path [draw,->,alice step] (fig-Multi-User Cycles-d-x) to[star] (-2em, 0cm |- fig-Disconnection-a);
\end{tikzpicture}
}
\label{fig:Disconnection-a}
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "\vWrapPlus" [> "\eWrapPlus"] -> {
   "\vSimpleTimes" [> "\eNestedBob"'],
   "\vMultiCycleTimes" [> "\eMultiCycleAliceTimes"]
  }
 }
};
\end{tikzpicture}
\end{subfigure}
%%%%%%%%
\hfill
%%%%%%%%
\begin{subfigure}[t]{0.21\textwidth}
\centering
\caption{
\begin{tikzpicture}[remember picture, overlay]
\path (0cm,0cm) node (fig-Disconnection-b) [anchor=base] {\strut};
\path [draw,->,alice step] (fig-Disconnection-a) to[star] (-2em, 0cm |- fig-Disconnection-b);
\end{tikzpicture}
}
\label{fig:Disconnection-b}
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "\vWrapPlus" [> "\eWrapPlus"] -> {
   "\vSimpleTimes" [> "\eNestedBob"'] -> {
    {},
    "\vMultiCycleTimes" [> "\eDisconnectionAlice", >alice edge]
   },
   {}
  }
 }
};
\end{tikzpicture}
\end{subfigure}
%%%%%%%%
\hfill
%%%%%%%%
\begin{subfigure}[t]{0.21\textwidth}
\centering
\caption{
\begin{tikzpicture}[remember picture, overlay]
\path (0cm,0cm) node (fig-Disconnection-c) [anchor=base] {\strut};
\path [draw,->,bob step] (fig-Disconnection-a) to[out=15,in=165,star] (-2em, 0cm |- fig-Disconnection-c);
\end{tikzpicture}
}
\label{fig:Disconnection-c}
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "\vWrapPlus" [> "\eWrapPlus"] -> {
   {},
   "\vMultiCycleTimes" [> "\eMultiCycleAliceTimes"] -> {
    "\vSimpleTimes" [> "\eDisconnectionBob"', >bob edge],
    {}
   }
  }
 }
};
\end{tikzpicture}
\end{subfigure}
%%%%%%%%
\hfill
%%%%%%%%
\begin{subfigure}[t]{0.21\textwidth}
\centering
\caption{
\begin{tikzpicture}[remember picture, overlay]
\path (0cm,0cm) node (fig-Disconnection-d) [anchor=base] {\strut};
\path [draw,->,merge step] (fig-Disconnection-b) to[out=15,in=165] (-2em, 0cm |- fig-Disconnection-d);
\path [draw,->,merge step] (fig-Disconnection-c) to (-2em, 0cm |- fig-Disconnection-d);
\end{tikzpicture}
}
\label{fig:Disconnection-d}
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "\vWrapPlus" [> "\eWrapPlus"]
 }
};
\path (-0.5cm,-2cm) graph[graph style] { plus/"\vSimpleTimes" };
\path ( 0.5cm,-2cm) graph[graph style] { times/"\vMultiCycleTimes" };
\path[draw,->,merge edge] (times) to ["\eMultiCycleBobTimes"'anchor=north,out=-135,in=-45] (plus);
\path[draw,->,merge edge] (plus) to ["\eMultiCycleBobPlus"anchor=south,out=45,in=135] (times);
\end{tikzpicture}
\end{subfigure}
%%%%%%%%
\hfill{}
\caption{Example of Disconnection.}
\label{fig:Disconnection}
\end{figure*}
}

@@

This section introduces collaborative structure editing in Grove.
We begin in \autoref{sub:Program Representation} by covering how we use graphs to represent expressions with holes,
\autoref{sub:Single-User Actions} explores actions performed by a single user, Alice,
and \autoref{sub:Multi-user Interactions} explores actions performed by multiple users, Alice and Bob,
each of whom is editing their own instance of a program.

@@

Program state is a directed graph with a distinguished root vertex.
%For example, \autoref{fig:A simple tree with a hole-b} shows the graph corresponding to the term in \autoref{fig:A simple tree with a hole-a}.
Each vertex represents a term and is labeled with both a globally unique identifier and a term constructor.
In \autoref{fig:A simple tree with a hole-b}, \vSimpleTimes{} and \vSimpleX{} have the constructors~\texttt{*} and \texttt{var(x)}, respectively. (For the sake of compactness, in our figures, we abbreviate \texttt{var(x)} as simply \texttt{x}.)
The root vertex has the distinguished identifier~0 and the constructor~\textbullet.

@@

@@

To start our examples, Alice moves her cursor to the hole in~\texttt{x * \hole} in \autoref{fig:A simple tree with a hole-b}
and constructs the variable~\texttt{y} as shown in \autoref{fig:Construction-b}.
This action corresponds to creating \eConstructionY{} from \vSimpleX{} at the \texttt{R} position to a newly
created vertex, \vConstructionY{}, containing the variable reference~\texttt{y}.
The resulting graph, shown in \autoref{fig:Construction-b}, represents the expression \texttt{x * y}.

@@

If a user wants to reposition code from one place to another, we have two options.
The first is to delete the code from its old position, and
reconstruct the code in its new position.
However, edges and vertices have unique identifiers in Grove 
to support the collaborative editing features we discuss next,
so reconstruction will not preserve identity of vertices and edges. 
Consequently, we take the second option: explicit repositioning actions that preserve identity.
These actions would be triggered by, for example, a drag and drop, or a cut and paste
(but not copy and paste, or a second paste after a cut, which do not need to
preserve identity and so would rely on reconstruction as already described in \autoref{sub:Construction}).

In terms of the graph, a repositioning action involves simply deleting the edge
from the original parent then adding an edge from the new parent.
Thus the code is actually \textit{repositioned}, not copied.

For example, in \autoref{fig:Move} Alice continues by
repositioning \texttt{\hole{} * y} from the left child of \texttt{+} to its right child.
This corresponds to the following sequence of graph edits, shown in \autoref{fig:Move}:
(a) delete \eWrapPlus{} leaving \vSimpleTimes{} temporarily orphaned and 
(b) add \eMoveTimes{} to \vSimpleTimes{} in the right child position of \vWrapPlus{}.

@@

Having described single-user actions,
we turn our attention to how Grove behaves in the presence of multiple users. 
Everything we discuss in this section generalizes to any number of users,
but for simplicity we consider only two: Alice and Bob. 
Alice and Bob each maintain their own editor instance, and perform 
actions relative to its state (using their own cursors, which we discuss in \autoref{sub:Cursors}).
We start each user where we left off in the previous section, i.e., in the state shown in \autoref{fig:Move-b}.

\todo{START REV HERE}

@@

Alice adds~\vDifferentAlice{} as the left child of \vSimpleTimes{}
and Bob changes \vConstructionY{} to \vDifferentBob{}.
Before transmitting their edits to each other,
Alice thus has the graph in \autoref{fig:Editing Different Parts of the Code-a}
and Bob has the graph in \autoref{fig:Editing Different Parts of the Code-b}.

%\figureCommutativity
\begin{figure}
\centering
\begin{tikzpicture}
\path (-3cm, 0cm) node (a) [align=center]       {Original \\ Version};
\path ( 0cm, 1cm) node (b) [align=center,alice node] {Alice's  \\ Version};
\path ( 0cm,-1cm) node (c) [align=center,bob node]   {Bob's    \\ Version};
\path ( 3cm, 0cm) node (d) [align=center]       {Synchronized \\ Version};
\path [draw,->,alice step] (a) -- node [pos=0.7,auto,align=center] {Alice's \\ Edits} (b);
\path [draw,->,bob step]   (a) -- node [pos=0.7,auto,swap,align=center]      {Bob's \\ Edits}   (c);
\path [draw,->,merge step] (b) -- node [pos=0.3,auto] {Sync} (d);
\path [draw,->,merge step] (c) -- node [pos=0.3,auto,swap]      {Sync} (d);
\end{tikzpicture}
\caption{Commutativity of Edits.}
\label{fig:Commutativity}
\end{figure}

@@

@@

After the edits in the previous subsection synchronize,
Alice changes~\vDifferentAlice{} to~\vNestedAlice{}, which results in the graph in \autoref{fig:Editing Nested Parts of the Code-a}.
On the other hand, Bob moves~\vSimpleTimes{} and its children from the left child of~\vWrapPlus{} to its right child
by deleting~\eMoveTimes{} and adding~\eNestedBob{}, which results in the graph in \autoref{fig:Editing Nested Parts of the Code-b}.
Since vertex and edge identifiers are preserved by these,
both Alice's and Bob's edits can be applied to the other's graph,
which results in both Alice and Bob having the graph in \autoref{fig:Editing Nested Parts of the Code-c}.

A consequence of this design is that we allow users to modify
parts of the code that have been deleted.  For example,
Alice might receive Bob's deletion of~\eMoveTimes{} before making her
changes and before receiving Bob's addition of~\eNestedBob{}.
Her edits would then be within the code deleted by Bob.
TODO: She can restore that code using a restoration action add a fresh edge to it.

@@

Finally, we address our first kind of conflict between edits.
Alice adds~\vMultiChildAlice{} as the right child of~\vWrapPlus{}
while Bob adds~\vMultiChildBob{} also as the right child of~\vWrapPlus{}.
These result in the graphs in \autoref{fig:Multi-child-a} and \autoref{fig:Multi-child-b}, respectively.
In particular pay attention to the edges going out of the \texttt{R} position of \vWrapPlus{}.
In \autoref{fig:Multi-child-a} it is~\eMultiChildAlice{}, which points to~\vMultiChildAlice{},
and in \autoref{fig:Multi-child-b} it is~\eMultiChildBob{}, which points to~\vMultiChildBob{}.

We merge these by including \emph{both} edges as edges going out of the \texttt{R} position
of \vWrapPlus{}.
This results in \autoref{fig:Multi-child-c}.
The conflict between Alice's and Bob's changes is
represented by the fact that there are multiple edges going out of one index (i.e., the \texttt{R}
position of \vWrapPlus{}).
In the graph representation, we flag this conflict by highlighting its edges
in red (TODO).
In the textual representation, we flag this conflict with the notation~\texttt{\{ x | y \}}.
This is a conflict that must eventually be resolved by the user by deleting
one of the edges and editing the other to be the merged value.
For example, one could delete \eMultiChildBob{} and then change~\vMultiChildAlice{}
to be~\texttt{x * y}.
Note that users can continue to edit the code while in this conflicted state.
It just must be resolved before the the code is run.\footnote{It
may be possible to develop evaluation models that allow these sorts of conflicts,
but that is beyond the scope of this paper.}

@@

An example of this is shown in \autoref{fig:Multi-parent}.
\autoref{fig:Multi-parent-a} shows the graph after Alice and Bob have made some more edits,
and Alice and Bob's editors are synchronized with each other.
This graph represents the expression~\texttt{(w + \hole{}) + \hole{})}.
Alice moves the variable~\texttt{w} to the left-most hole,
while Bob moves it to the right-most hole.
This results in the graphs in \autoref{fig:Multi-parent-b} and \autoref{fig:Multi-parent-c},
which represent respectively the expression~\texttt{(\hole{} + w) + \hole{}} for Alice
and the expression~\texttt{(\hole{} + \hole{}) + w} for Bob.
In both cases, these edits are achieved with one edge delete and one edge add.
Alice deletes~\eNestedAlice{} and adds~\eMultiParentAlice{}, while
Bob deletes~\eNestedAlice{} and adds~\eMultiParentBob{}.
As explained in Subsection~TODO:REF, deletes are idempotent so Alice and Bob both deleting~\eNestedAlice{} is not a problem.
However, both~\eMultiParentAlice{} (added by Alice) and~\eMultiParentBob{} (added by Bob) point to the same vertex.
This generates \autoref{fig:Multi-parent-d}, in which~\vNestedAlice{} has two parents pointing to it.
We call this situation a multi-parent conflict.

@@

\subsubsection{Cycles}
\label{sub:Multi-User Cycles}

@@

For example, consider the situation in \autoref{fig:Multi-User Cycles}.
\autoref{fig:Multi-User Cycles-a} shows the graph after Alice and Bob have made some more edits,
and Alice and Bob's editors are synchronized with each other.
This graph represents the expression~\texttt{((\hole+\hole)+(\hole*\hole))+\hole}.
Alice moves~\vMultiCycleTimes{} to the \texttt{R} child of~\vWrapPlus{}
and then~\vMultiCyclePlus{} underneath.
This results in \autoref{fig:Multi-User Cycles-b},
which represents the expression~\texttt{(\hole+\hole)+((\hole*\hole)+\hole)}.
On the other hand, Bob does the same move
but puts~\vMultiCycleTimes{} and under~\vMultiCyclePlus{}.
This results in \autoref{fig:Multi-User Cycles-c},
which represents the expression~\texttt{(\hole+\hole)+((\hole+\hole)*\hole)}.

On their own, neither of these edits creates a cycle.
However, merging the edit actions of both Alice and Bob results in the graph
in \autoref{fig:Multi-User Cycles-d}, which has a cycle
between~\vMultiCycleTimes{} and~\vMultiCyclePlus{}.
This situation is akin to a merge conflict in traditional version control systems.
Resolving it requires user input, but is as simple as
either deleting~\eMultiCycleBobPlus{} and~\eMultiCycleBobTimes{}~(thus favoring Alice's version)
or deleting~\eMultiCycleAliceTimes{} and~\eMultiCycleAlicePlus{}~(thus favoring Bob's version).
Since a cycle like this that is connected to the root vertex always
contains vertexes with multiple parents that if removed would break the cycle,
we use the user interface described in \autoref{sub:Multi-parent conflicts}
to display these kinds of programs.
For example, \autoref{fig:Multi-User Cycles-d} would be displayed
as shown in TODO:FIGURE.
We leave the user interface considerations of resolving the conflict such a
cycle represents to future work.

@@

For example, consider the situation in \autoref{fig:Disconnection}.
\autoref{fig:Disconnection-a} shows the graph after Alice and Bob have made some more edits,
and Alice and Bob's editors are synchronized with each other.
This graph represents the expression~\texttt{(\hole*\hole)+(\hole*\hole)}.
Alice moves~\vMultiCycleTimes{} under~\vSimpleTimes{}.
This results in \autoref{fig:Disconnection-b},
which represents the expression~\texttt{(\hole*(\hole*\hole))+\hole}.
On the other hand, Bob puts~\vSimpleTimes{} and under~\vMultiCycleTimes{}.
This results in \autoref{fig:Disconnection-c},
which represents the expression~\texttt{\hole+((\hole*\hole)*\hole)}.

When these edits are merged, the result is the graph in \autoref{fig:Disconnection-d}.
Since Alice deleted~\eNestedBob{}
and Bob deleted~\eMultiCycleAliceTimes{},
when these are merged~\vMultiCycleTimes{} and~\vSimpleTimes{}
are disconnected from the root.
